Particle Smoothing Algorithms for Variable Rate Models

### 2012-09-13 response_to_reviewers.tex

\documentclass{article}

\usepackage{amsmath}
\usepackage{IEEEtrantools}
\usepackage{color}

\newenvironment{meta}[0]{\color{red} \em}{}

\begin{document}

{\meta Many thanks for the comments, folks. I've made changes to address all but one, with which I have a bit of a philosophical disagreement. Reviewer 3's comments were particularly useful and prompted me to rethink carefully the purpose of the simulations section. Responses are inline with the individual points below. }

\section*{Reviewer Comments}

\subsection*{Reviewer: 1}

Recommendation: AQ - Publish In Minor, Required Changes

\subsubsection*{Comments}
This is an interesting, well written paper  that suggests new sampling strategies for smoothing  in variable rate models. The proposed method  rejuvenates the variable rate filtering particles that usually have poor smoothing performance due to the loss of particle diversity particularly when state discontinuities occur.

The paper relays on the key idea, using the previously generated filtering particles as a proposal  to the target conditional distributions, which is inherited from the previous work of the author [Godsill et al. 2004] . The proposed algorithms apply a backward pass by sampling the  target conditional distribution $P( \theta^{-}_{n} |  . )$  for $n=N…1$.   Two methodologies are proposed for sampling the target conditional distribution of the  variable rate structured "conditionally linear Gaussian models"  and "conditionally deterministic models". The former samples from a discrete set of re-weighted filtering particles according to well known Rao-Blackwellization approach and the other uses an  MCMC sampling scheme where the proposals are constructed by a  modified  filtering particle set.

Both algorithms are sophisticated techniques for rejuvenating the particles and can be  interpreted as a way of reconstructing the particle history by using the filtering particles sampled at each filtering step. Reference on the convergence property of the algorithms and a  discussion to understand the practical aspects of the the algorithm (Ex. whether is it  possible to increase the online filtering performance by using retrospective smoothing steps, the ability of the algorithm to  find the untracked modes of the posterior - this can also be suported with simulation results )  would be useful  for the reader in order to understand the limits of the algorithms.

{\meta I've added a brief note on convergence, which is inherited from the basic results for the forward-filtering-backward-sampling algorithm. I've also added an explicit example in the simulations of the smoother representing multi-modality where some particle contain a jump and others don't. Fixed interval smoothing won't help online filtering. Fixed-lag smoothing would, and this is covered elsewhere, especially in [Whitely et al. 2011]. }

In the simulations section, smoothing performance of the algorithms are compared to filtering and  the filtering- smoothing results.  In both tests improvement is achieved in means of RMSE. I see that in the  simulations filter and filter- smoother is able to represent the solution to the data. However it would be much more interesting and convincing to see an illustration of a scenario  where the filter is not able to represent the solution due to discarded particles in the resampling steps  (Ex.  poor estimate for the changepoint location or  parameters,  misses a jump or unable to track the maneuver due to multi-modality)  but  the smoother is able track the solution.

{\meta Because the smoother uses filter particles as a proposal, if the filter fails catastrophically (e.g. track loss) then the smoother will also fail. I have added an example where the multi-modality of the particle representation is lost due to resampling in the filter-smoother, but not the filter. }

Minor comments:

Page: 37 line 36 :  " $\theta$ " should be  "  $\theta'$ "
Fig: 4 step 11 " $\theta*$ " should be  " $\theta'*$  "

{\meta Well spotted. Now correct.}

\subsection*{Reviewer: 2}

Recommendation: A - Publish Unaltered

\subsubsection*{Comments}
(There are no comments. Please check to see if comments were included as a file attachment with this e-mail or as an attachment in your Author Center.)

{\meta No file attached, so I assume there are genuinely no comments. }

\subsection*{Reviewer: 3}

Recommendation: RQ - Review Again After Major Changes

\subsubsection*{Comments}
The authors propose a smoothing PF for variable rate models. The proposed method is essentially an adaptation of existing methods for fixed-rate models to the variable rate model. I feel there is enough substance in the adaptation process to merit publication provided some revisions are performed. My main concerns are with the performance analysis:

- It is never clarified but I'm assuming that the smoothing results are for fixed-interval smoothing, i.e., estimates at time $t_n, n=1,...,N$ use measurements from times $t_1,...,t_N$. This should be clearly stated.

{\meta I feel this is clear from the introduction and the maths, but it's now explicitly stated as well. }

- Only 10 realisations are used! It's difficult to draw conclusions with such a small sample size.

{\meta Ok. Now 100.}

- In Figs 7 and 10 the effective sample size would be a better measure of performance than the number of unique particles.

- The principal improvement offered by the more sophisticated particle smoothers compared to the filter-smoother is increased particle diversity. This should improve estimation performance but the improvements in consistency may be even greater. This is alluded to in the paper and partially covered by Figs 6 and 9 which show plots of sample trajectories. However, a more rigorous consistency analysis should be performed. For example, a simple approach would be to construct a Gaussian approximation to the posterior, establish a confidence ellipsoid and then count the number of times the true value falls within this ellipsoid.

{\meta These two comments concern the same question: How should particle approximations of a distribution be compared? There is, of course, much more to a distributional estimate than how close its mean is to the true value, so we do need something more than just the RMSE, as you suggest. Ideally, we would like to compare our set of particles to the true posterior distribution, using statistics such as the effective sample size (ESS), or normalised estimation error squared (NEES). However this true posterior is not available. As a substitute, either a Gaussian approximation or a particle approximation with loads of particles are sometimes used as ``truth''. This is far from ideal, and I find the latter particularly unsettling, because we end up testing our particle algorithms only against their own best performance.

My preferred alternative for measuring consistency is a mean ``empirical NEES'' statistic which takes a value between 0 and 1 and takes into account the estimation error, covariance estimate and diversity of the particle approximation. I've added details and plots of this metric to the paper. I think that together with the RMSE values, this provides sufficient evidence of the improvement provided by the smoother. }

- Comparing the proposed method and the filter-smoother for equal sample sizes doesn't really make sense given that the former has $O(N^2)$ expense while the latter has $O(N)$ expense. The filter-smoother should be implemented with a much larger sample size such that both algorithms have approximately equal expense. This approach, which is taken in (Fearnhead et al, 2010), seems much fairer.

{\meta I completely disagree. I think this would be misleading and is unnecessary for demonstrating the purpose of the new smoothers.

For a start, the filter-smoother and the smoother provide estimates with different qualities, and comparing them like this would be to compare apples with oranges. The principal reason for using a particle smoother is to rejuvenate a particle population with low path-space diversity. This is demonstrated by the increased number of unique particles and the improved consistency. (The reduced RMSE is a consequent bonus.) In contrast, [Fearnhead et al. 2010] are comparing a variety of smoothing algorithms all of which do the same thing, produce a diverse set of marginal smoothing particles. Their methods are not appropriate here.

Furthermore, if we were trying to solve a particular practical problem, such as producing the best state estimator possible for an offline tracker, then it would be a sensible test. However, the example simulations here are entirely fictional, cooked up simply to demonstrate that the new smoothing algorithm functions as expected. The problem is that whatever this test demonstrated about relative performance, the result could be reversed by changing the set-up of the simulation. }

- Results should be shown for a range of sample sizes so we can get an idea of the required sample size for (close to) optimal performance.

{\meta Yes. I've added graphs showing the effect of varying the number of smoothing trajectories for the tracking case. The finance results are similar and so omitted for space. }

Some minor comments:
- A Rao-Blackwellised VRPF was also developed in (Morelande and Gordon, 2009). Their formulation was more general than that used here since it permits the model to be partially linear-Gaussian.

{\meta Yes. This should be and now is referenced. Actually, their formulation is not more general, just different. The Rao-Blackwellisation in our paper is of the state variables when the dynamics are linear-Gaussian, while [Morelande and Gordon, 2009] Rao-Blackwellise the linear parts of the motion parameters, but in a conditionally deterministic model. We could extend our work by using their Rao-Blackwellisation scheme on the piecewise-deterministic smoother. }

- Recently, O(N) schemes for particle smoothing which improve upon the filter-smoother have been proposed by (Fearnhead et al, 2010) and (Douc et al, 2011). Perhaps these could be used for VR filtering. They should at least be mentioned.

{\meta I've added a reference to [Douc et al., 2011]. This is similar to our Metropolis-Hastings-based method which is already referenced (but unhelpfully so as it is still awaiting publication). ([Bunch and Godsill, 2012]). The Fearnhead method is for marginal (as opposed to joint) smoothing distributions only, so it doesn't really help. I've mentioned this in section IV. }

- My understanding is that in (33) $\tau_{K_{n-1}+1}$ should be drawn from a truncated distribution which forces it to be greater than $t_{n-1}$. According to the notation of (4), shouldn't the denominator then be $S(\tau_{K_{n-1}},...)$ rather than $S(\tau_{K_{n-1}+1},...)$?

{\meta Yes. Well spotted! Corrected. }

- On page 13, just before (51), there is a $\tilde{\theta}$ which should be a $\theta$.

{\meta Fixed. }

- In Fig. 9(a), are these results for the filter-smoother?

{\meta Yes. Clarified.}

- I'm not sure all the appendices are necessary since the results can be found in other, easily obtained sources. At least

{\meta I agree. I've cut the first two, giving the derivation and initialisation of the two-filter smoother. The discretisation of the finance model remains, as I can't find a complete and equivalent version in a published work. }

\end{document}

### 2012-09-13 revisions_VRsmoothing.tex
Revisions

In a variable rate setting, changepoints and associated parameters are modelled as a marked point process (MPP), the mathematical properties of which are explored comprehensively in \cite{Jacobsen2006}. Conditional upon this MPP the state evolves according to some benign dynamics. In \cite{Godsill2007,Whiteley2011}, the conditional state evolution is treated as deterministic, while in \cite{Godsill2007a,Christensen2012} a conditionally linear-Gaussian state model is considered.

@@

The VRPF allows the changepoint sequence --- and hence the current state --- to be estimated sequentially as observations are received. However, estimates can often be improved later once further observations have been made. In this paper, we address the problem of smoothing in variable rate models, i.e. the estimation of the changepoint sequence and latent state given all the observations. We achieve this with an efficient backward sweep through the observations, in a new version of the method for standard HMMs described in \cite{Godsill2004}. Two new schemes are introduced: one for conditionally linear-Gaussian models which exploits the method of Rao-Blackwellisation; and a second for use with conditionally-deterministic models which uses an augmented target distribution in the style of an SMC sampler \cite{DelMoral2006}.

@@

Target tracking algorithms are commonly based upon fixed rate models (see, e.g. \cite{Li2003} for a thorough survey), in which the target kinematics (position, velocity, etc.) are estimated at a set of fixed times at which observations (e.g. radar measurements) are made. In \cite{Godsill2007a,Godsill2007,Whiteley2011,Bunch2012a}, variable rate models were introduced for tracking. The state trajectory is divided up by a set of changepoints, between which the motion follows a deterministic path governed by motion parameters (accelerations, etc.) which are fixed for that division. In this paper we use the 2-dimensional intrinsic coordinate model of \cite{Bunch2012a} to test the new smoothing algorithm.

@@

The transition term, $p(\theta_{n \setminus n-1} | \theta_{n-1}^-)$, has a similar form to the changepoint prior of (\ref{eq:cp_sequence_prior}) \cite{Jacobsen2006}, but an additional survivor function is included to account for the condition that changepoints cannot occur before $t_{n-1}$,
%
\begin{IEEEeqnarray}{rCl}
\IEEEeqnarraymulticol{3}{l}{p(\theta_{n \setminus n-1} | \theta_{n-1}^-)} \nonumber \\
  &=& S(\tau_{K_n},u_{K_n}, t_n) / S(\tau_{K_{n-1}+1},u_{K_{n-1}+1}, t_{n-1}) \nonumber \\
  & & \times \prod\limits_{j:t_{n-1} \leq \tau_j < t_n} p(\tau_j, u_j| \tau_{j-1}, u_{j-1})  \label{eq:cp_sequence_trandens}     .
\end{IEEEeqnarray}
%\begin{IEEEeqnarray}{rCl}
%\IEEEeqnarraymulticol{3}{l}{p(\theta_{n \setminus n-1} | \theta_{n-1})} \nonumber \\
%    & = & p(\tau_{K_{n-1}+1:K_n}, u_{K_{n-1}+1:K_n}, \tau_{K_n+1}>t_n|\tau_{K_{n-1}+1}>t_{n-1}, \tau_{K_{n-1}}, u_{K_{n-1}}) \nonumber \\
%    & = & \begin{cases} S(\tau_{K_n}, t_n) \prod\limits_{j:t_{n-1} \leq \tau_j < t_n} p(\tau_j, u_j| \tau_{j-1}, u_{j-1}, \tau_j>t_{n-1}) & K_n > K_{n-1} \\
%                        S(\tau_{K_n}, t_n) / S(\tau_{K_n}, t_{n-1}) & K_n = K_{n-1} \end{cases} \IEEEeqnarraynumspace \label{eq:cp_sequence_trandens}
%\end{IEEEeqnarray}

@@

@@

The target distribution for a variable rate smoothing algorithm is the posterior distribution over the entire changepoint sequence, $p(\theta|y_{1:N})$. This distribution is approximated by the final step of the VRPF. However, in the same manner as the fixed rate filter-smoother of \cite{Kitagawa1996}, this approximation is likely to lack path-space diversity; because of the necessary resampling step in the filtering algorithm, the particles all share the same set of changepoints before a particular time, with variation only appearing for changepoints closer to $T$. For a better characterisation of the smoothing distribution, it is necessary to rejuvenate the set of particles. This is achieved here with a backward pass through the observations in an extension of the forward-backward method described in \cite{Godsill2004}.

@@

The second state-dependent term in (\ref{eq:rb-vrps_back_cond_weight}) is the likelihood, $p(y_{n+1:N}|x_n, \tilde{\theta}_{n}^+)$. This may be regarded as an improper density over $x_n$, and may be calculated analytically using a backwards Kalman filter, in a similar manner to that used in the two-filter smoother \cite{Fraser1969,Kitagawa1994,Sarkka2012,Gelb1974}. Such a backwards Kalman filter uses the following recursions, which are derived in appendix~\ref{app:backwards_KF},
%
\begin{IEEEeqnarray}{rCl}
 p(y_{n+1:N}|x_n, \theta_{n}^+) & \propto & \mathcal{N}(x_n|\tilde{m}_n^-, \tilde{P}_n^-) \\
 p(y_{n:N}|x_n, \theta_{n}^+) & \propto & \mathcal{N}(x_n|\tilde{m}_n, \tilde{P}_n)
\end{IEEEeqnarray}
%
\begin{IEEEeqnarray}{rCl}
 \tilde{m}_n^- & = & A_{n+1}^{-1} \tilde{m}_{n+1} \nonumber \\
 \tilde{P}_n^- & = & A_{n+1}^{-1} (\tilde{P}_{n+1} + Q_{n+1}) A_{n+1}^{-T} \label{eq:backward_kf_predict} \\
 \nonumber \\
 \tilde{\mu}_n & = & C_n \tilde{m}_n^- \nonumber \\
 \tilde{S}_n   & = & C_n \tilde{P}_n^- C_n^T + R_n \nonumber \\
 \tilde{G}_n   & = & \tilde{P}_n^- C_n^T \tilde{S}_n^{-1} \nonumber \\
 \tilde{m}_n   & = & \tilde{m}_n^- + \tilde{G}_n (y_n - \tilde{\mu}_n) \nonumber \\
 \tilde{P}_n   & = & \tilde{P}_n^- - \tilde{G}_n \tilde{S}_n \tilde{G}_n^T \label{eq:backward_kf_update}     .
\end{IEEEeqnarray}

Exact methods for initialising this recursion are discussed in appendix~\ref{app:init_backwards_KF}, and also in \cite{Kitagawa1994}. However, it is often sufficient to use an approximation, for example $\tilde{m}_N^{(i)} \gets m_N^{(i)}$ and $\tilde{P}_N^{(i)} \gets P_N^{(i)}$.

@@

The algorithmic complexity of this variable rate particle smoother for conditionally linear-Gaussian models is $O(\aleph_F \times \aleph_S \times N)$. It is possible to reduce this by using an MCMC sampling scheme in the style of \cite{Bunch2012}, which avoids the necessity of calculating the sampling weights for every filter particle.

@@

The augmented target distribution may be sampled using Markov chain Monte Carlo (MCMC) \cite{Gilks1996}, in a similar manner to the fixed rate smoother of \cite{Bunch2012}. Metropolis-Hastings (MH) steps are conducted to draw samples, $\{\tilde{\theta}_{n}^-, u_{K_n}\}$, from the target distribution by sampling the proposal and accepting the new values with a given probability. Note that IS could be used instead of MCMC, but this would require resampling steps, which would lead to a loss of path-space diversity similar to that suffered by the particle filter.

@@

\begin{figure}
\fbox{\parbox{\columnwidth}{
\begin{algorithmic}[1]
\linespread{1.5} \selectfont
  \STATE Run variable rate particle filter to approximate $p(\tilde{\theta}_{n}^-|y_{1:n})$ with weighted particles $\{\tilde{\theta}_{n}^{-(i)}, w_{n}^{(i)}\}$. Store all results.
  \FOR{$i=1 \dots \aleph_S$}
  	\STATE Initialise particle using $\tilde{\theta}^{(i)} \sim \sum_j w_N^{(j)} \delta_{\tilde{\theta}^{(j)}}(\tilde{\theta})$.
    \FOR{$n=N \dots 1$}
      \STATE Initialise chain with $\tilde{\theta}_n^{-(i)(0)} \gets \tilde{\theta}_n^{-(i)}$.
      \FOR{$m=1 \dots M$}
        \STATE Propose a new history $\tilde{\theta}_n^{-(i)*} \sim \sum_j w_n^{(j)} \delta_{\tilde{\theta}_{n}^{-(j)}}(\tilde{\theta}_{n}^-)$.
        \STATE Propose a parameter $u_{K_n}'^* \sim q(u_{K_n}|\tilde{\theta}_{n}^-, \tilde{\theta}_{n}^{+}, y_{1:N})$.
        \STATE Replace $u_{K_n}^*$ with $u_{K_n}'^*$ to form $\tilde{\theta}_n^{-'(i)*}$.
        \STATE Calculate $\alpha_n^{(m)}$ using (\ref{eq:pd-vrps_tpr}) and (\ref{eq:pd-vrps_ap}).
        \STATE With probability $\alpha_n^{(m)}$, $\tilde{\theta}_n^{-(i)(m)} \gets \tilde{\theta}_n^{-(i)*}$. \\ Otherwise, $\tilde{\theta}_n^{-(i)(m)} \gets \tilde{\theta}_n^{-(i)(m-1)}$.
      \ENDFOR
      \STATE Store final sample $\tilde{\theta}_n^{-(i)} \gets \tilde{\theta}_n^{-(i)(M)}$.
    \ENDFOR
  \ENDFOR
\end{algorithmic}
}}
\caption{MCMC piecewise-deterministic variable rate particle smoother}
\label{alg:VRPS}
\end{figure}

The number of MH steps, $M$, at each observation time, $n$, allows a trade-off of performance against time. Larger values of $M$ will ensure more unique samples in the smoothing approximation, but will also take longer to execute. See \cite{Bunch2012} for analysis of this trade-off on a fixed rate model. In the simulations discussed in this paper, $M=1$ was used throughout.

@@

Comparisons of accuracy are most easily conducted by comparing state estimates. These can be calculated by simply taking a mean of the particle values (using the Gaussian means for the linear-Gaussian case). The accuracy is then assessed by comparing the root mean squared error (RMSE) of the resulting estimators.

@@

In addition, the benefit of employing a smoothing algorithm is that it ensures a more diverse particle representation of the posterior state distribution. To demonstrate this effect, the number of unique state samples is plotted over time in the following examples.

@@

\begin{figure}[!t]
\centering
\includegraphics[width=0.9\columnwidth]{example_data.pdf}
\caption{An example simulated data set, showing value (top) and trend (bottom). Value observations are overlayed. Jump times are shown as dotted vertical lines.}
\label{fig:example_data}
\end{figure}

\begin{figure}[!t]
\centering
\subfloat[]{\includegraphics[width=0.9\columnwidth]{example_filtersmoother_state.pdf}} \\
\subfloat[]{\includegraphics[width=0.9\columnwidth]{example_smoother_state.pdf}}
\caption{Trend estimates from (a) the filter-smoother and (b) the VRPS. Solid lines show means of each particle. Dashed lines show mean $\pm$2 standard deviations.}
\label{fig:example_state}
\end{figure}

Performance was evaluated by testing on 10 scenarios, each with 1000 observations. At each observation time, an estimate of the value and trend is made by taking the average of Gaussian means generated by an RTS smoother for each particle. The resulting root mean square errors (RMSEs) are displayed in table~\ref{tab:finance_state_performance}, demonstrating an improvement in accuracy from the smoother.

\begin{table}%
\caption{State estimation performance.}
\label{tab:finance_state_performance}
\centering
\renewcommand{\arraystretch}{1.5}
\begin{tabular}{|c|c|c|c|}
\hline
 &\raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{1.5cm}{\centering VRPF}} & \raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{1.5cm}{\centering Filter-smoother}} & \raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{1.5cm}{\centering VRPS}} \\
\hline \hline
\raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{2cm}{\centering mean value estimate RMSE}}   & $5.31 \times 10^{-4}$ & $4.48 \times 10^{-4}$ & $4.16 \times 10^{-4}$ \\
\hline
\raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{2cm}{\centering mean trend estimate RMSE}}   & $2.56 \times 10^{-2}$ & $1.79 \times 10^{-2}$ & $1.49 \times 10^{-2}$ \\
\hline
\end{tabular}
\end{table}

>>>>

In a variable rate setting, changepoints and associated parameters are modelled as a marked point process (MPP), the mathematical properties of which are explored comprehensively in \cite{Jacobsen2006}. Conditional upon this MPP the state evolves according to some benign dynamics. In \cite{Godsill2007,Whiteley2011,Morelande2009a}, the conditional state evolution is treated as deterministic, while in \cite{Godsill2007a,Christensen2012} a conditionally linear-Gaussian state model is considered.

@@

The VRPF allows the changepoint sequence --- and hence the current state --- to be estimated sequentially as observations are received. However, estimates can often be improved later once further observations have been made. In this paper, we address the problem of fixed-interval smoothing in variable rate models, i.e. the estimation of the changepoint sequence and latent state given all the observations. We achieve this with an efficient backward sweep through the observations, in a new version of the method for standard HMMs described in \cite{Godsill2004}. Two new schemes are introduced: one for conditionally linear-Gaussian models which exploits the method of Rao-Blackwellisation; and a second for use with conditionally-deterministic models which uses an augmented target distribution in the style of an SMC sampler \cite{DelMoral2006}.

@@

Target tracking algorithms are commonly based upon fixed rate models (see, e.g. \cite{Li2003} for a thorough survey), in which the target kinematics (position, velocity, etc.) are estimated at a set of fixed times at which observations (e.g. radar measurements) are made. In \cite{Godsill2007a,Godsill2007,Whiteley2011,Morelande2009a,Bunch2012a}, variable rate models were introduced for tracking. The state trajectory is divided up by a set of changepoints, between which the motion follows a deterministic path governed by motion parameters (accelerations, etc.) which are fixed for that division. In this paper we use the 2-dimensional intrinsic coordinate model of \cite{Bunch2012a} to test the new smoothing algorithm.

@@

The transition term, $p(\theta_{n \setminus n-1} | \theta_{n-1}^-)$, has a similar form to the changepoint prior of (\ref{eq:cp_sequence_prior}) \cite{Jacobsen2006}, but an additional survivor function is included to account for the condition that changepoints cannot occur before $t_{n-1}$,
%
\begin{IEEEeqnarray}{rCl}
\IEEEeqnarraymulticol{3}{l}{p(\theta_{n \setminus n-1} | \theta_{n-1}^-)} \nonumber \\
  &=& S(\tau_{K_n},u_{K_n}, t_n) / S(\tau_{K_{n-1}},u_{K_{n-1}}, t_{n-1}) \nonumber \\
  & & \times \prod\limits_{j:t_{n-1} \leq \tau_j < t_n} p(\tau_j, u_j| \tau_{j-1}, u_{j-1})  \label{eq:cp_sequence_trandens}     .
\end{IEEEeqnarray}
%\begin{IEEEeqnarray}{rCl}
%\IEEEeqnarraymulticol{3}{l}{p(\theta_{n \setminus n-1} | \theta_{n-1})} \nonumber \\
%    & = & p(\tau_{K_{n-1}+1:K_n}, u_{K_{n-1}+1:K_n}, \tau_{K_n+1}>t_n|\tau_{K_{n-1}+1}>t_{n-1}, \tau_{K_{n-1}}, u_{K_{n-1}}) \nonumber \\
%    & = & \begin{cases} S(\tau_{K_n}, t_n) \prod\limits_{j:t_{n-1} \leq \tau_j < t_n} p(\tau_j, u_j| \tau_{j-1}, u_{j-1}, \tau_j>t_{n-1}) & K_n > K_{n-1} \\
%                        S(\tau_{K_n}, t_n) / S(\tau_{K_n}, t_{n-1}) & K_n = K_{n-1} \end{cases} \IEEEeqnarraynumspace \label{eq:cp_sequence_trandens}
%\end{IEEEeqnarray}

@@

In \cite{Morelande2009a}, an improvement to this algorithm for conditionally-deterministic models was introduced in which the linear parts of the motion parameters are Rao-Blackwellised. Note that this is different to the scheme used in the previous section in which the state variables are Rao-Blackwellised.

@@

The target distribution for a variable rate smoothing algorithm is the posterior distribution over the entire changepoint sequence, $p(\theta|y_{1:N})$. This distribution is approximated by the final step of the VRPF. However, in the same manner as the fixed rate filter-smoother of \cite{Kitagawa1996}, this approximation is likely to lack path-space diversity; because of the necessary resampling step in the filtering algorithm, the particles all share the same set of changepoints before a particular time, with variation only appearing for changepoints closer to $T$. For a better characterisation of the smoothing distribution, it is necessary to rejuvenate the set of particles. This is achieved here with a backward pass through the observations in an extension of the forward-filtering-backward-sampling (FFBS) method described in \cite{Godsill2004}. Since these algorithms are based on the FFBS algorithm, the convergence proofs of \cite{Godsill2004,Douc2011} are believed to apply.

For standard HMMs, it is common to distinguish between joint smoothing distributions, $p(x_{1:N}|y_{1:N})$, and marginal smoothing distributions, $p(x_k|y_{1:N})$. Note that for variable rate models, the latter is ill defined --- we need the entire sequence of changepoints --- and thus algorithms for estimating marginal smoothing distributions, such as reweighting smoothers \cite{Doucet2000} and two-filter smoothers \cite{Briers2010,Fearnhead2010} will not be considered here.

@@

The second state-dependent term in (\ref{eq:rb-vrps_back_cond_weight}) is the likelihood, $p(y_{n+1:N}|x_n, \theta_{n}^+)$. This may be regarded as an improper density over $x_n$, and may be calculated analytically using a backwards Kalman filter, in a similar manner to that used in the two-filter smoother \cite{Fraser1969,Kitagawa1994,Sarkka2012,Gelb1974,Sarkka2012}. Such a backwards Kalman filter uses the following recursions,% which are derived in appendix~\ref{app:backwards_KF},
%
\begin{IEEEeqnarray}{rCl}
 p(y_{n+1:N}|x_n, \theta_{n}^+) & \propto & \mathcal{N}(x_n|\tilde{m}_n^-, \tilde{P}_n^-) \\
 p(y_{n:N}|x_n, \theta_{n}^+) & \propto & \mathcal{N}(x_n|\tilde{m}_n, \tilde{P}_n)
\end{IEEEeqnarray}
%
\begin{IEEEeqnarray}{rCl}
 \tilde{m}_n^- & = & A_{n+1}^{-1} \tilde{m}_{n+1} \nonumber \\
 \tilde{P}_n^- & = & A_{n+1}^{-1} (\tilde{P}_{n+1} + Q_{n+1}) A_{n+1}^{-T} \label{eq:backward_kf_predict} \\
 \nonumber \\
 \tilde{\mu}_n & = & C_n \tilde{m}_n^- \nonumber \\
 \tilde{S}_n   & = & C_n \tilde{P}_n^- C_n^T + R_n \nonumber \\
 \tilde{G}_n   & = & \tilde{P}_n^- C_n^T \tilde{S}_n^{-1} \nonumber \\
 \tilde{m}_n   & = & \tilde{m}_n^- + \tilde{G}_n (y_n - \tilde{\mu}_n) \nonumber \\
 \tilde{P}_n   & = & \tilde{P}_n^- - \tilde{G}_n \tilde{S}_n \tilde{G}_n^T \label{eq:backward_kf_update}     .
\end{IEEEeqnarray}

Exact methods for initialising this recursion are discussed in \cite{Kitagawa1994,Sarkka2012}. However, it is often sufficient to use an approximation, for example $\tilde{m}_N^{(i)} \gets m_N^{(i)}$ and $\tilde{P}_N^{(i)} \gets P_N^{(i)}$. %appendix~\ref{app:init_backwards_KF}, and also in

@@

The algorithmic complexity of this variable rate particle smoother for conditionally linear-Gaussian models is $O(\aleph_F \times \aleph_S \times N)$. It is possible to reduce this by using a rejection sampling \cite{Douc2011} or MCMC \cite{Bunch2012} sampling scheme, which avoids the necessity of calculating the sampling weights for every filter particle.

@@

The augmented target distribution may be sampled using Markov chain Monte Carlo (MCMC) \cite{Gilks1996}, in a similar manner to the fixed rate smoother of \cite{Bunch2012}. Metropolis-Hastings (MH) steps are conducted to draw samples, $\{\tilde{\theta}_{n}^{-'}, u_{K_n}\}$, from the target distribution by sampling the proposal and accepting the new values with a given probability. Note that IS could be used instead of MCMC, but this would require resampling steps, which would lead to a loss of path-space diversity similar to that suffered by the particle filter.

@@

\begin{figure}
\fbox{\parbox{\columnwidth}{
\begin{algorithmic}[1]
\linespread{1.5} \selectfont
  \STATE Run variable rate particle filter to approximate $p(\tilde{\theta}_{n}^-|y_{1:n})$ with weighted particles $\{\tilde{\theta}_{n}^{-(i)}, w_{n}^{(i)}\}$. Store all results.
  \FOR{$i=1 \dots \aleph_S$}
  	\STATE Initialise particle using $\tilde{\theta}^{(i)} \sim \sum_j w_N^{(j)} \delta_{\tilde{\theta}^{(j)}}(\tilde{\theta})$.
    \FOR{$n=N \dots 1$}
      \STATE Initialise chain with $\tilde{\theta}_n^{-(i)(0)} \gets \tilde{\theta}_n^{-(i)}$.
      \FOR{$m=1 \dots M$}
        \STATE Propose a new history $\tilde{\theta}_n^{-(i)*} \sim \sum_j w_n^{(j)} \delta_{\tilde{\theta}_{n}^{-(j)}}(\tilde{\theta}_{n}^-)$.
        \STATE Propose a parameter $u_{K_n}'^* \sim q(u_{K_n}|\tilde{\theta}_{n}^-, \tilde{\theta}_{n}^{+}, y_{1:N})$.
        \STATE Replace $u_{K_n}^*$ with $u_{K_n}'^*$ to form $\tilde{\theta}_n^{-'(i)*}$.
        \STATE Calculate $\alpha_n^{(m)}$ using (\ref{eq:pd-vrps_tpr}) and (\ref{eq:pd-vrps_ap}).
        \STATE With probability $\alpha_n^{(m)}$, $\tilde{\theta}_n^{-(i)(m)} \gets \tilde{\theta}_n^{-'(i)*}$. \\ Otherwise, $\tilde{\theta}_n^{-(i)(m)} \gets \tilde{\theta}_n^{-(i)(m-1)}$.
      \ENDFOR
      \STATE Store final sample $\tilde{\theta}_n^{-(i)} \gets \tilde{\theta}_n^{-(i)(M)}$.
    \ENDFOR
  \ENDFOR
\end{algorithmic}
}}
\caption{MCMC piecewise-deterministic variable rate particle smoother}
\label{alg:VRPS}
\end{figure}

The number of MH steps, $M$, at each observation time, $n$, allows a trade-off of performance against time. Larger values of $M$ will ensure more unique samples in the smoothing approximation, but will also take longer to execute. See \cite{Bunch2012} for analysis of this trade-off on a fixed rate model. The resulting computational complexity of the algorithm is $O(M \times \aleph_S \times N)$.

@@

Comparisons of accuracy are most easily conducted by comparing state estimates. These can be calculated by simply taking a mean of the particle values (using the Gaussian means for the linear-Gaussian case). The accuracy is then assessed by comparing the size of estimation errors of the resulting estimators. In addition, the benefit of employing a smoothing algorithm is that it ensures a more diverse particle representation of the posterior state distribution. To demonstrate this effect, a measure of estimator consistency is calculated by which to compare the algorithms, the mean empirical normalised estimation error squared (MENEES). This is defined as,  %the number of unique state samples is plotted over time in the following examples. A
%
\begin{IEEEeqnarray}{rCl}
 \text{MENEES}_k & = & (\hat{x}_k - x_k^*)^T \hat{P}_k (\hat{x}_k - x_k^*) \nonumber \\ 
 \hat{x}_k & = & \sum_i x_{k}^{(i)} \nonumber \\
 \hat{P}_k & = & \sum_i (x_{k}^{(i)} - x_k^*) (x_{k}^{(i)} - x_k^*)^T \nonumber     ,
\end{IEEEeqnarray}

where $x_k^*$ is the true state, and $x_k^{(i)}$ are the individual particle states. This statistic takes a value between $0$ and $1$, where $1$ is scored by a degenerate particle estimate (fewer linearly independent particles than state dimensions). For a non-degenerate estimate, smaller values correspond to smaller estimation errors relative to the estimated covariance. Thus, it incorporates the diversity of the particle set as well the accuracy of the estimate and covariance.

@@

@@

%
\begin{figure}[!t]
\centering
\includegraphics[width=0.9\columnwidth]{example_data.pdf}
\caption{An example simulated data set, showing value (top) and trend (bottom). Value observations are overlayed. Jump times are shown as dotted vertical lines.}
\label{fig:example_data}
\end{figure}
%
\begin{figure}[!t]
\centering
\subfloat[]{\includegraphics[width=0.9\columnwidth]{example_filtersmoother_state.pdf}} \\
\subfloat[]{\includegraphics[width=0.9\columnwidth]{example_smoother_state.pdf}}
\caption{Trend estimates from (a) the filter-smoother and (b) smoother. Solid lines show means of each particle. Dashed lines show mean $\pm$2 standard deviations.}
\label{fig:example_state}
\end{figure}
%
Performance was evaluated by testing on 100 scenarios, each with 1000 observations. At each observation time, an estimate of the value and trend is made by taking the average of Gaussian means generated by an RTS smoother for each particle. The resulting mean absolute error over all scenarios is shown in figure~\ref{fig:finance_rmse}, demonstrating a small improvement in accuracy from the smoother. More significantly, MENEES is shown in figure~\ref{fig:finance_menees}, showing a similar consistency between filter and smoother. The filter-smoother produces an MENEES of $1$ at almost all time instants, because there is only one unique particle in the set.% are displayed in table~\ref{tab:finance_state_performance}
%
\begin{figure}[!t]
\centering
\subfloat[]{\includegraphics[width=0.9\columnwidth]{finance_value_rmse.pdf}} \\
\subfloat[]{\includegraphics[width=0.9\columnwidth]{finance_trend_rmse.pdf}}
\caption{Value (a) and trend (b) mean absolute error over 100 scenarios using filter, filter-smoother, and smoother.}
\label{fig:finance_rmse}
\end{figure}
%
\begin{figure}[!t]
\centering
\includegraphics[width=0.9\columnwidth]{finance_menees.pdf}
\caption{MENEES over 100 scenarios using filter, filter-smoother, and smoother.}
\label{fig:finance_menees}
\end{figure}
%
%\begin{table}%
%\caption{State estimation performance.}
%\label{tab:finance_state_performance}
%\centering
%\renewcommand{\arraystretch}{1.5}
%\begin{tabular}{|c|c|c|c|}
